Renewal and stability in populations structured by remaining years of life

### 2016-01-22 ThanoRepro/ReplyReviewers.tex

\documentclass{article}

\usepackage{amsmath}
\usepackage{natbib}
\bibpunct{(}{)}{,}{a}{}{;} 
\newcommand{\dd}{\; \mathrm{d}}
\usepackage{url}

\begin{document}
\title{Reply to Reviewers of submission 3043: ``Renewal and stability in populations structured by remaining years of
life''}
\author{Author Redacted}
\maketitle

I thank reviewers A and B for taking the time to engage with and comment on this
submission, and the assistant editor for his suggestions and guidance, and to
all involved for a swift first round of review. I've made an attempt to increase
appeal by reducing unneeded sentences and statements, and by changing
language, for instance swapping out thanatological age for years left in most
cases. Reviewers comments are then treated individually.

However, I first wish to politely state that I think DR is a good home for
purely theoretical or mathematical investigations because (in general and in no
order) a) DR has no space limitations, b) readers of DR come in almost
exclusively via content-based internet searches rather than periodical
readership (ergo there is no need to profile readership), c) the alternatives
(e.g., Theoretical Population Biology and Mathematical Population Studies) are not open access and so I have ethical objections to
them, and d) DR already sponsors a formal relationships series (ergo formal demography is demonstrably in-universe). I
should like to think that DR covers the full breadth of researchers that
demographers do. This article is standard fare for formal demography, and formal
demography is the foundation of the discipline. Browsing through the titles of
recently publishes articles in DR, I'm lead to believe that appealing to a broad
audience is not itself a criterion for publication, and I do not believe that it
should be either.

That said, I've followed advice to make the article more palatable and
approachable, and these offset somewhat the space saved by following through on
reviewer A's wish for the article to be more compact.

Comments in order:

Editor:

\end{document}

### 2016-01-22 ThanoRepro/ThanoRepro.tex
starting to make changes for first round of review. started draft of response to reviewers

%\setlength\PreviewBorder{5pt}
\usepackage{natbib}
\bibpunct{(}{)}{,}{a}{}{;} 
\usepackage{url}
\usepackage{nth}
% for the d in integrals
\newcommand{\dd}{\; \mathrm{d}}
\newcommand{\ec}{\quad\quad\text{,}}
\newcommand{\ep}{\quad\quad\text{.}}

\usepackage[top=2in, bottom=1.5in, left=1in, right=1in]{geometry}
\usepackage{setspace}
\doublespacing

@@

\subsection*{Background}
The Lotka-Leslie renewal model is the core of formal demography. This model is
structured by chronological age, and it does not account for thanatological age.

\subsection*{Objective}
I derive a specification of the classic renewal equation that is structured by
thanatological age rather than by chronological age. I give both continuous
and discrete variants of the derived model, and relate these to the Lotka-Leslie
renewal model.

\subsection*{Results}
In stability, the thanatological and chronological renewal models are
commensurable, implying identical intrinsic growth rates.
I demonstrate approximate symmetry between chronological and
thanatological age structure in stability when subject to
intrinsic growth rates equal magnitude and opposite sign.

@@

\subsection*{Contribution}
The thanatological renewal model offers a new perspective on population renewal,
and it is valid more generally as an aspect of birth-death processes.
\vspace{2cm}

\section*{Introduction}

Contemporary demography is built upon a small set of empirical
regularities with respect to chronological age patterns. The models and methods
of demography are in varying degrees prefaced on such regularity. There is evidence that many demographic phenomena are best described in the
aggregatate as a function of time since birth, and others of time until death,
while others can be a function of both to some degree
\citep{riffe2015ttd,wolf2015disability}. This observation motivates
incorporating this aspect time-to-death into formal demographic methods and
models.

In this paper, I explore some formal demographic consequences of a particular redefinition of age. Instead of counting age as the time passed since birth, consider age as the amount of time left until death. The first
kind of age is called chronological age, and the second kind of age is called
thanatological age.\footnote{This quantity is referred to as \textit{residual} lifetime
in the reliability literaure, or just time-to-death (TTD) elsewhere.} Individuals in this case move in the same direction along
imaginary life lines, but the reference point for thanatological age is at
the end of the life line instead of at the beginning. For individuals, the timing of events in life is exactly identified using either of the two definitions of age, but
aggregate patterns differ. 
%Temporal perspectives and rescalings of the lifecourse are abundant in
% demography already. Such work has enriched the discipline by producing new
% insights, questions, and conundrums.
% Fertility and reproduction are plausibly a function of both age perspectives,
 % and the question of how to best partition them over time is not investigated here.

The renewal model relates
fertility, mortality, and population structure in a parsimonious form, revealing
the long term consequences of a particular set of conditions. All diverse lines
of contemporary demographic research are relatable to the renewal model, and by
extention to stable population theory. The purpose of this paper is to describe
the Lotka-Leslie renewal model when structured by thanatological age.
%Some might say that stable population
%theory is not relevant to everyday demography, since populations are always
% changing. Even if change precludes the notion of stability as a prediction, models of long-run consequnces still give a
%useful measure of the demographic force implied by a given set of rates.
%In this and other ways, stability is never far away. That population
% projections lead into a stable state means that some form of stability is implied by the routine practice of demography.
%Stability or even stationarity are imaginable for the world population of
% humans one day. Models of stationary populations are easily applied by analogy to
%closed or nearly-closed birth cohorts. Finally, stable population theory is
%relavant for modelling some non-human populations. 
%We have stated that many lifecourse transitions covary more with remaining
%lifetime than with life lived. In an even more general sense, by thinking about
% remaining lifetime as a particular class of time-to-transition, one may generalize the lifetable methods presented here into new area applications, such as education, partnering, the economic lifecycle, or migration.
To present the renewal model based on thanatological age is not necessarily a
suggestion for a better way to model or project population. Rather, the
thanatological renewal model represents an alternative and coherent way to
describe the process of population renewal by relating renewal to a death flow
rather than to a birth flow. I have described a limited case of this
perspective on renewal for the case of stationary populations
\citep{riffe2015force}, and here provide a broader treatment allowing for
growth in the population.
The chronological-age classified renewal model and the thanatological renewal
model both describe the movement of population stocks over time, but from opposite vantage points. %Specifying a model of thanatological renewal expands the toolbox available to approach complex temporal population structure.

I couch the present inquiry in terms of aggregate renewal of human populations,
but the renewal model is more generally a description of the birth-death
process \citep{cox1962,feller1941integral}. As such, the thanatological
renewal model may be applicable in myriad other areas, such as queueing,
infectious disease processes, or aspects of renewable aggregates in general.
Given the breadth of birth-death models and the speculative applicability of the thanatological renewal model, I am
content for the present inquiry to establish the validity of the model. Perhaps
there exists a birth-death process for which modelling as a death-birth process
makes the most sense. In order to establish the \textit{soundness} of this model as a description of the human renewal process
would require considerable empirical groundwork--- It would imply demonstrating
that the thanatological age pattern of fertility is empirically regular, and
that the time-to-death structure of the population is empirically regular.
Such an endeavor would be misplaced, as the objective of this inquiry is not to
improve the practice of empirical demography. Instead the objective is to deepen
our understanding of the Lotka-Leslie renewal model. The chronological and
thanatological renewal models imply one another in the state of stability. We
therefore present this model as a property implied by the Lotka-Leslie renewal
model, as two sides of the same coin. The result is however more general than
that of literal birth and death in human populations.

The next section of this article describes the lifetable method of transforming age-classified
data into thanatological age classes. I then derive the continuous version of
the thanatological renewal model, and some other stable properties. This is
followed by the matrix version of the model, which is explained alongside the
lifecycle graph. Finally, I explore patterns in stable age structure compared
between the two models. I conclude that thanatological renewal offers
a new perspective on the familiar Lotka-Leslie model.

\section*{Requisites of a thanatological renewal model}
The Lotka-Leslie renewal model \citep{leslie1945use,sharpe1911problem} is
situated in a unisex homogenous population framework. The model expressed in continous mathematics contains a
component for survival, $\ell(a)$, a scaling factor for growth, $e^{-ra}$, and a
component for fertility, $m(a)$:

@@

Each of these elements is structured by
chronological age, $a$. The model can be approximated by observed data because
chronological age is typically known. The survival function, $\ell(a)$, derives
from a lifetable, usually based on age-specific event-exposure mortality rates,
which require counts of decedents and an estimate of the
population exposed to risk in the same chronological age classes.
Chronological age in the growth factor, $e^{-ra}$ pertains to the passage of
time more than it does to chronological age (these are synchronous for birth
cohorts). The product of this growth factor and $\ell(a)$ gives a
function proportional to the stable age structure, $c(a)$, and it is often
easier to imagine these two elements together as one for this reason. Female
fertility rates, $m(a)$, are based on birth counts classified by age-of-mother, and the same variety of exposure to risk. For the two essential rate componenents underlying the renewal equation~\eqref{eq:lotka}, classification by chronological age is essentially given--- it is deterministic in the sense that time passed since birth is typically known by individuals. The kinds of errors in such classification are mostly due to innumeracy, rounding, exaggeration, and
similar kinds of biases, but even in such cases, chronological age is
fundamentally knowable.

Thanatological age is different in this respect because we do not know when we
will die--- it is only knowable retrospectively or probabilistically. Under a set of simple assumptions, the same lifetable used to
estimate $\ell(a)$ can be used to
partition and recombine counts observed by chronological age into estimates
classified by thanatological age. Our objective is then to estimate an age
structure component, similar to $\ell(a)e^{-ra}$, and a fertility component
simialr to $m(a)$, but each for the thanatological case. Taken together, these
componenents will yield a thanatological renewal model.

\subsection*{A lifetable approximation of thanatological age}
%Direct information on the timing of demographic transitions with respect to
%death is available retrospectively from linked population registers or
%panel surveys with mortality follow-ups.
%Some surveys also ask respondents in various ways about their perceived future
% lifetime, which may underly those demographic transitions governed to some degree by volition.

In general, to make a lifetable-based projective statement about the remaining
time distribution of a current stock of population, it is best to to account for population subgroups
with different mortality and to account for uncertainty about future changes in
mortality.
The present inquiry approximates thanatological age using lifetable-projected remaining lifetime,
which is intended to approximate remaining lifetime under the strong
assumptions of homogenous populations und fixed vital rates. 
%Chronological age-classified counts are redistributed over future death times
% according to a fixed mortality schedule.
%In general, a lifetable approach to transforming chronologically classified
%data into thanatologically classified data is sound if the given lifetable
%accurately describes the future mortality of the age-classified population, and
%if the chronological and thantological age patterns are independent.
Under these assumptions, each birth cohort undergoes the same
pattern of attrition. In a period setting, this means that we can distribute the
remaining lifespans of the population in each age class using the same
lifetable.
%This is the case we explore, and so the following formulas assume a single lifetable pattern.

The basic transformation to thanatological age proceeds by defining a
conditional deaths density function, $f(y|a)$. For chronological
age zero, $f(y|0)$ is identical to $d(a)$, the lifetable deaths distribution
(based on a unity radix).
For ages higher than 0, simply condition on survival:\footnote{In the reliability literature, $f(y|a)$ is
denoted by $ X - t | X > t $, where $X$ is the total lifespan and $t$ is the
current chronological age attained.}
\begin{equation}
\label{eq:vaupel1}
f(y | a) = \mu(a+y)\frac{\ell(a+y)}{\ell(a)} \ec
\end{equation}
where $\mu(a)$ is the force of mortality.

A convenient discretization of \eqref{eq:vaupel1} is to simply work with the
$d_a$ column of the lifetable (here with single ages):
\begin{equation}
\label{eq:dyadiscret}
f_{y, a} =  \frac{d_{a+y}+d_{a+y+1}}{L_a + L_{a+1}} \ec
\end{equation}
where $L_a$ is lifetable exposure.\footnote{Equation~\eqref{eq:dyadiscret}
will incur some error depending on the method used to derive $L_a$. To remove
this error, rescale over the $a$ margin so that each distribution conditioned on
$a$ sums to 1. Working with $\ell_a$ rather than $L_a$ eliminates error and
yields similar results, but is not necessarily better.} One produces in this way a probability density function over future death times for each chronological age,
$a$. The $f(y|a)$- weighted average of $y$ for a given $a$ returns the familiar
remaining life expectancy column of the lifetable, $e(a)$. Since $1 = \int_0^\infty f(y|a) \dd y$ for each $a$,
this density is useful for decomposing data classified by chronological age,
such as the population pyramid, into thanatological age classes. If $P(a)$ is a
chronological age-classified population count, we can derive $P(y)$, a
thanatological age-classified population count using\footnote{A
different-looking but identical definition is found in
\citet{brouard1989mouvements}.}

@@

Figure~\ref{fig:USdecomp} shows this exercise for 2010 US data \citep{HMD},
assuming the 2010 period mortality schedule remains constant. When looking from the chronological age perspective, this decomposition reveals thanatological
heterogeneity, but when viewed from the thanatological perspective
(Figure~\ref{fig:USdecomp}) it reveals chronological age heterogeneity and the
overall thanatological age profile, the future death flow of the current
population.
Figure~\ref{fig:USrecomp} shows this recomposition for the same 2010
US data. The outer profile of Figure~\ref{fig:USrecomp} is $P(y)$, under our assumptions.

% both figures created in R/Pyramids.R
\begin{figure}[ht!]
	\caption{2010 US population structure}
	\begin{center}
	\begin{subfigure}{.45\textwidth}
		\caption{Chronological age-structure (years lived) decomposed by
		thanatological age (years left).}
		\label{fig:USdecomp}
		\includegraphics[scale=.6]{Figures/US2010Age.pdf}
	\end{subfigure}
	~
	\begin{subfigure}{.45\textwidth}
		\caption{Thanatological age-structure (years left) decomposed by chronological
		age (years lived).}
		\label{fig:USrecomp}
		\includegraphics[scale=.6]{Figures/US2010Thano.pdf}
	\end{subfigure}
	\\
	\small{*Population and mortality data from the HMD.}
	\end{center}
\end{figure}

@@

\section*{The thanatological projection matrix}
These descriptions can be made more explicit, and in some ways more tractable,
by hashing out the projection matrix that corresponds to the thanatological renewal model. As with the
chronological age-structured Leslie matrix, the thanatological projection
matrix, $\textbf{Y}$, is square and of dimension $y \times y$, where $y$ is the number
of equal-interval thanatological age classes into which the population is
structured.
The matrix contains elements for survival and elements for fertility. Unlike
Leslie matrices, $\textbf{Y}$ is not sparse, but is populated primarily
with positive entries greater than zero. In the
following, I illustrate using a 6$\times$6 matrix.

Mortality occurs in only the population class with zero
remaining years of life. The population of thanatological age 1 in year $t$
moves to 0 in year $t + 1$ and dies in the course of that year. Thus, rather than in the
subdiagonal, survival elements are located in the superdiagonal. All survival
values are 1, since there is no decrement until after living through age 0. The survival component of $\textbf{Y}$ is organized as in
Matrix~\ref{matrix:Ysurv}. To imagine a lifecycle graph, each age decrements by
one through each lower age in successive order until being absorbed by death
after completing thanatological age zero.

\begin{matrix}[h!]
\centering
\caption{Survival component of unisex thanatological projection matrix,
$\textbf{Y}$}
\label{matrix:Ysurv}
$\bordermatrix{{e_y } & 0_t & 1_t & 2_t & 3_t & 4_t & 5_t\cr 
                0_{t+1} & 0    &  1   & 0    & 0    & 0    & 0   \cr
                1_{t+1} & 0    &  0   & 1    & 0    & 0    & 0   \cr 
                2_{t+1} & 0    &  0   & 0    & 1    & 0    & 0   \cr 
                3_{t+1} & 0    &  0   & 0    & 0    & 1    & 0   \cr 
                4_{t+1} & 0    &  0   & 0    & 0    & 0    & 1   \cr
                5_{t+1} & 0    &  0   & 0    & 0    & 0    & 0   }$
\end{matrix}

 Fertility inputs to the matrix are derived from single-sex thanatological
 fertility and the lifetable $d_a$ distribution, where $a$ indexes
 chronological age and is equated to $y$ remaining years of life for members
 just born. It is simpler to imagine $d_a$ in this case as indexing the
 future lifespan distribution of newborns \citep{riffe2015force}. Fertility in a
 thanatologically structured human population occurs in all but the very highest
 thanatological age classes, which can be thought of as structural zeros for
 simplicity, as they only contain pre-menarchical females that will have very long lives.

@@

 Further, infant mortality, $\gamma_y \cdot d_0$, located in the first row, must also be discounted, since part
 of the mortality will occur in the same year $t$ and the rest in year $t + 1$. 
 The first row of fertility must be further discounted by a factor, $\lambda$,
 in order to account for the fact that infant mortality is higher in the lower Lexis 
 triangle than in the upper. Of those infants who die in the first year of life,
 a proportion equal to $\lambda$ do not survive to December \nth{31} of the
 calendar year in which they were born.\footnote{$\lambda$ can be derived directly from death counts
 data classified by Lexis triangles. If the demographer
 does not have information to derive $\lambda$ directly, ad hoc or semidirect
 methods may be used to assign a reasonable proportion, such as 0.9 for
 contemporary low mortality populations.} The fertility component of $\textbf{Y}$ is then composed as in Matrix~\ref{matrix:Yfert}.

\begin{matrix}[h!]
\centering
\caption{Fertility component of unisex thanatological projection matrix,
$\textbf{Y}$}
\label{matrix:Yfert}
$\bordermatrix{
  {e_y } \vspace{.6em}&                0_t  & 1_t  & 2_t  & 3_t  & 4_t  & 5_t\cr 
   0_{t+1} \vspace{.6em}& (1-\lambda) \tfrac{\gamma_0d_0}{2} & (1-\lambda) \gamma_1d_0 & (1-\lambda)
   \gamma_2d_0 & (1-\lambda) \gamma_3d_0 & (1-\lambda) \gamma_4d_0 & 0 \cr 
   1_{t+1} \vspace{.6em}& \tfrac{\gamma_0d_1}{2} & \gamma_1d_1 & \gamma_2d_1 & \gamma_3d_1 & \gamma_4d_1
   & 0   \cr 2_{t+1} \vspace{.6em}& \tfrac{\gamma_0d_2}{2} & \gamma_1d_2 & \gamma_2d_2 & \gamma_3d_2 & \gamma_4d_2
   & 0   \cr 3_{t+1} \vspace{.6em}& \tfrac{\gamma_0d_3}{2} & \gamma_1d_3 & \gamma_2d_3 & \gamma_3d_3 & \gamma_4d_3
   & 0   \cr 4_{t+1} \vspace{.6em}& \tfrac{\gamma_0d_4}{2} & \gamma_1d_4 & \gamma_2d_4 & \gamma_3d_4 & \gamma_4d_4
   & 0   \cr 5_{t+1} \vspace{.6em}& \tfrac{\gamma_0d_5}{2} & \gamma_1d_5 & \gamma_2d_5 & \gamma_3d_5 & \gamma_4d_5
   & 0   }$
\end{matrix}
\FloatBarrier

The survival and fertility components of $\textbf{Y}$ add together elementwise,
thus the full 6$\times$6 matrix is composed as in Matrix~\ref{matrix:Y}.

>>>>

%\setlength\PreviewBorder{5pt}
\usepackage{natbib}
\bibpunct{(}{)}{,}{a}{}{;} 
\usepackage{url}
\usepackage{nth}
% for the d in integrals
\newcommand{\dd}{\; \mathrm{d}}
\newcommand{\ec}{\quad\quad\text{,}}
\newcommand{\ep}{\quad\quad\text{.}}
\defcitealias{HMD}{HMD}
\usepackage[top=2in, bottom=1.5in, left=1in, right=1in]{geometry}
\usepackage{setspace}
\doublespacing

@@

\subsection*{Background}
The Lotka-Leslie renewal model is the core of formal demography. This model is
structured by age, and it does not account for time until death.

\subsection*{Objective}
I derive a specification of the classic renewal equation that is structured by
years left rather than by years lived. I give both continuous
and discrete variants of the derived model, and relate these to the Lotka-Leslie
renewal model.

\subsection*{Results}
In stability, the years lived and left renewal models are
commensurable, implying identical intrinsic growth rates.
I demonstrate approximate symmetry between years lived and left age structure in
stability when subject to intrinsic growth rates equal magnitude and opposite sign.

@@

\subsection*{Contribution}
The years left renewal model offers a new perspective on population renewal,
and it is valid more generally as an aspect of birth-death processes.
\vspace{2cm}

\section*{Introduction}
Contemporary demography is built upon a small set of empirically regular age
patterns. The models and methods
of demography are in varying degrees prefaced on such regularity. There is evidence that many demographic phenomena are best described in the
aggregatate as a function of time since birth, and others of time until death,
while others can be a function of both to some degree
\citep{riffe2015ttd,wolf2015disability}. Further, aggregate indices
motivated by measures of remaining lifetime have also become widely-used
indicators of ageing \citep{sanderson2007new}. These observations motivate
incorporating time-to-death into formal demographic methods and models.

In this paper, I explore some formal demographic consequences of a particular
redefinition of age. Instead of counting age as the time passed since birth,
consider the amount of time left until death.\footnote{Elsewhere this quantity
is referred to as \textit{residual} lifetime, time-to-death (TTD), prospective
age \citep{sanderson2007new}, or thanatological age \citep{riffe2015force}.}
Individuals in this case move in the same direction along imaginary life lines,
but the reference point for measurement is switched from birth to death. For
individuals, the timing of events in life is exactly identified using either of
years lived or left, but aggregate patterns differ.
%Temporal perspectives and rescalings of the lifecourse are abundant in
% demography already. Such work has enriched the discipline by producing new
% insights, questions, and conundrums.
% Fertility and reproduction are plausibly a function of both age perspectives,
 % and the question of how to best partition them over time is not investigated here.

The renewal model relates
fertility, mortality, and population structure in a parsimonious form, revealing
the long term consequences of a set of vital rates. All lines
of contemporary demographic research are relatable to the renewal model, and by
extention to stable population theory. The purpose of this paper is to describe
the Lotka-Leslie renewal model when structured by years of life left.
%Some might say that stable population
%theory is not relevant to everyday demography, since populations are always
% changing. Even if change precludes the notion of stability as a prediction, models of long-run consequnces still give a
%useful measure of the demographic force implied by a given set of rates.
%In this and other ways, stability is never far away. That population
% projections lead into a stable state means that some form of stability is implied by the routine practice of demography.
%Stability or even stationarity are imaginable for the world population of
% humans one day. Models of stationary populations are easily applied by analogy to
%closed or nearly-closed birth cohorts. Finally, stable population theory is
%relavant for modelling some non-human populations. 
%We have stated that many lifecourse transitions covary more with remaining
%lifetime than with life lived. In an even more general sense, by thinking about
% remaining lifetime as a particular class of time-to-transition, one may generalize the lifetable methods presented here into new area applications, such as education, partnering, the economic lifecycle, or migration.
%To present the renewal model based on thanatological age is not necessarily a
%suggestion for a better way to model or project population. Rather, 
The years left renewal model represents an alternative and coherent way to
describe the process of population renewal by relating renewal to a death flow
rather than to a birth flow. I have described a limited case of this
perspective on renewal for the case of stationary populations
\citep{riffe2015force}, and here provide a broader treatment allowing for
growth in the population.
The years lived renewal model and the years left renewal
model both describe the movement of population stocks over time, but from opposite vantage points. %Specifying a model of thanatological renewal expands the toolbox available to approach complex temporal population structure.

I couch the present inquiry in terms of aggregate renewal of human populations,
but the renewal model is more generally a description of the birth-death
process \citep{cox1962,feller1941integral}, and so may be valid for other kinds
of renewable aggregates.
The present inquiry establishes the validity of the model, but does not argue
for its empirical practicality or soundness.
%Perhaps there exists a birth-death process for which modelling as a death-birth
% process makes the most sense. In order to establish the \textit{soundness} of this model as a description of the human renewal process
%would require considerable empirical groundwork--- It would imply demonstrating
%that the thanatological age pattern of fertility is empirically regular, and
%that the time-to-death structure of the population is empirically regular.
%Such an endeavor would be misplaced, as the objective of this inquiry is not to
%improve the practice of empirical demography. 
The years lived and left renewal models imply one another in the state of stability. We
therefore present the years left model as a property implied by the
standard Lotka-Leslie renewal model, as two sides of the same coin. The objective in this paper is to deepen
our understanding of the Lotka-Leslie renewal model, and to offer an extention
of stable population theory.

The next section of this article reviews the lifetable method of transforming
age-classified data into remaining years classes. I then derive the continuous version of
the years left renewal model, and some other stable properties. This is
followed by the matrix version of the model, which is explained alongside the
lifecycle graph. Finally, I explore patterns in stable age structure compared
between the two models. I conclude that years left renewal offers
a new perspective on the familiar Lotka-Leslie model.

\section*{Requisites of a years left renewal model}
The Lotka-Leslie renewal model \citep{leslie1945use,sharpe1911problem} is
situated in a closed, unisex, homogenous population framework. With
parameters equated, the continuous form of the model contains a component for
survival, $\ell(a)$, a scaling factor for growth, $e^{-ra}$, and a component for fertility, $m(a)$:

@@

Each of these elements is structured by
age, $a$. The model can be approximated by observed data because
age is typically known.% The survival function, $\ell(a)$, derives
%from a lifetable, usually based on age-specific event-exposure mortality rates,
%which require counts of decedents and an estimate of the
%population exposed to risk in the same chronological age classes.
Age in the growth factor, $e^{-ra}$, pertains to the passage of
time more than it does to chronological age (these are synchronous for birth
cohorts). The product of this growth factor and $\ell(a)$ gives a
function proportional to the stable age structure, $c(a)$, and it is often
intuitive to imagine these two elements together as one for this reason. Female
fertility rates, $m(a)$, are based on daughters born to mothers in this limited
single-sex model. This model lends itself to commonly collected data because age
is typically known, and can be asked.
% For the two essential rate componenents underlying the renewal
 % equation~\eqref{eq:lotka}, classification by chronological age is essentially given--- it is deterministic in the sense that time passed since birth is typically known by individuals. The kinds of errors in such classification are mostly due to innumeracy, rounding, exaggeration, and similar kinds of biases, but even in such cases, chronological age is fundamentally knowable.

Remaining lifetime is different in this respect because we do not know when we
will die--- it is only knowable retrospectively or probabilistically. Under a set of simple assumptions, the same lifetable used to
estimate $\ell(a)$ can be used to
partition and recombine counts observed by age into estimates
classified by years of life left. Our objective is then to estimate a population
structure component, similar to $\ell(a)e^{-ra}$, and a fertility component
similar to $m(a)$, but each for the years left case. Taken together, these
componenents yield the years left renewal model.

\subsection*{A lifetable approximation of life left}
%Direct information on the timing of demographic transitions with respect to
%death is available retrospectively from linked population registers or
%panel surveys with mortality follow-ups.
%Some surveys also ask respondents in various ways about their perceived future
% lifetime, which may underly those demographic transitions governed to some degree by volition.

%In general, to make a lifetable-based projective statement about the remaining
%time distribution of a current stock of population, it is best to to account
% for population subgroups with different mortality and to account for uncertainty about future changes in
%mortality.
The present inquiry approximates years left in the aggregate using
lifetable-projected remaining lifetime, which is intended to approximate remaining lifetime under the strong
assumptions of homogenous populations und fixed vital rates. 
%Chronological age-classified counts are redistributed over future death times
% according to a fixed mortality schedule.
%In general, a lifetable approach to transforming chronologically classified
%data into thanatologically classified data is sound if the given lifetable
%accurately describes the future mortality of the age-classified population, and
%if the chronological and thantological age patterns are independent.
Under these assumptions, each birth cohort undergoes the same
pattern of attrition. %In a period setting, this means that we can distribute
% the remaining lifespans of the population in each age class using the same
%lifetable.
%This is the case we explore, and so the following formulas assume a single lifetable pattern.

The basic transformation to years left starts with the well-known remaining
years density function, $f(y|a)$.
%For age zero, $f(y|0)$ is identical to $d(a)$, the lifetable deaths
% distribution (based on a unity radix).
%For ages higher than 0, simply condition on survival:
%\footnote{In the
% reliability literature, $f(y|a)$ is denoted by $ X - t | X > t $, where $X$ is the total lifespan and $t$ is the
%current chronological age attained.}
\begin{equation}
\label{eq:vaupel1}
f(y | a) = \mu(a+y)\frac{\ell(a+y)}{\ell(a)} \ec
\end{equation}
where $\mu(a)$ is the force of mortality.
%A convenient discretization of \eqref{eq:vaupel1} is to simply work with the
%$d_a$ column of the lifetable (here with single ages):
%\begin{equation}
%\label{eq:dyadiscret}
%f_{y, a} =  \frac{d_{a+y}+d_{a+y+1}}{L_a + L_{a+1}} \ec
%\end{equation}
%where $L_a$ is lifetable exposure.\footnote{Equation~\eqref{eq:dyadiscret}
%will incur some error depending on the method used to derive $L_a$. To remove
%this error, rescale over the $a$ margin so that each distribution conditioned
% on $a$ sums to 1. Working with $\ell_a$ rather than $L_a$ eliminates error and
%yields similar results, but is not necessarily better.} One produces in this
% way a probability density function over future death times for each chronological age, $a$. 
The $f(y|a)$- weighted average of $y$ for a given $a$ returns the remaining life
expectancy column of the lifetable, $e(a)$. $f(y|a)$ is also useful for
decomposing age-classified data, such as the
population pyramid, into remaining years classes. If $P(a)$ is an
age-classified population count, we can derive the years left version $P(y)$,
using\footnote{An identical definition is found in
\citet{brouard1989mouvements} and illustrated in \citet{brouard1986structure}.}

@@

Figure~\ref{fig:pop} shows this exercise for 2010 US data \citep{HMD},
assuming homogeneity and that the 2010 period mortality schedule is held fixed.
When looking from the age perspective (Figure~\ref{fig:USdecomp}), this
decomposition reveals projected years left heterogeneity within a stock of known age, but when viewed from the
years left perspective (Figure~\ref{fig:USrecomp}) it reveals age heterogeneity
and the projected overall years left profile--- the future death flow of the
current population, $P(y)$ under our assumptions.

% both figures created in R/Pyramids.R
\begin{figure}[ht!]
	\caption{2010 US population structure}
	\label{fig:pop}
	\begin{center}
	\begin{subfigure}{.45\textwidth}
		\caption{Age structure (years lived) decomposed by years left.}
		\label{fig:USdecomp}
		\includegraphics[scale=.6]{Figures/US2010Age.pdf}
	\end{subfigure}
	~
	\begin{subfigure}{.45\textwidth}
		\caption{Years left structure decomposed by years lived.}
		\label{fig:USrecomp}
		\includegraphics[scale=.6]{Figures/US2010Thano.pdf}
	\end{subfigure}
	\\
	\small{*Population and mortality data from the HMD.}
	\end{center}
\end{figure}

@@

\section*{The thanatological projection matrix}
These descriptions can be made more explicit, and in some ways more tractable,
by hashing out the projection matrix that corresponds to the thanatological renewal model. As with the
chronological age-structured Leslie matrix, the thanatological projection
matrix, $\textbf{Y}$, is square and of dimension $\omega \times \omega$, where $\omega$ is
the number of equal-interval thanatological age classes into which the population is
structured.
The matrix contains elements for survival and elements for fertility. Unlike
Leslie matrices, $\textbf{Y}$ is not sparse, but is populated primarily
with non-zero positive entries. In the
following, I illustrate using a 6$\times$6 matrix.

Mortality occurs in only the population class with zero
remaining years of life. The population of thanatological age 1 in year $t$
moves to 0 in year $t + 1$ and dies in the course of that year. Thus, rather than in the
subdiagonal, survival elements are located in the superdiagonal. All survival
values are 1, since there is no decrement until after living through age 0. The
survival component of $\textbf{Y}$, \textbf{S}, is organized as in
Matrix~\ref{matrix:Ysurv}. To imagine a lifecycle graph, each age decrements by one through each lower age in successive order until being absorbed by death
after completing thanatological age zero.

\begin{matrix}[h!]
\centering
\caption{Survival component,  \textbf{S}, of unisex thanatological projection
matrix.}
\label{matrix:Ysurv}
$\bordermatrix{{e_y } & 0_t & 1_t & 2_t & 3_t & 4_t & 5_t\cr 
                0_{t+1} & 0    &  1   & 0    & 0    & 0    & 0   \cr
                1_{t+1} & 0    &  0   & 1    & 0    & 0    & 0   \cr 
                2_{t+1} & 0    &  0   & 0    & 1    & 0    & 0   \cr 
                3_{t+1} & 0    &  0   & 0    & 0    & 1    & 0   \cr 
                4_{t+1} & 0    &  0   & 0    & 0    & 0    & 1   \cr
                5_{t+1} & 0    &  0   & 0    & 0    & 0    & 0   }$
\end{matrix}

 Fertility inputs to the matrix are derived from single-sex thanatological
 fertility and the lifetable $d_a$ distribution, where $a$ indexes
 chronological age and is equated to $y$ remaining years of life for members
 just born. It is simpler to imagine $d_a$ in this case as indexing the
 future lifespan distribution of newborns \citep{riffe2015force}. Fertility in a
 thanatologically structured human population occurs in all but the very highest
 thanatological age classes, which can be thought of as structural zeros, as they only contain pre-menarchical females that will have very long lives.

@@

 Further, infant mortality, $\gamma_y \cdot d_0$, located in the first row, must also be discounted, since part
 of the mortality will occur in the same year $t$ and the rest in year $t + 1$. 
 The first row of fertility must be further discounted by a factor, $\lambda$,
 in order to account for the fact that infant mortality is higher in the lower Lexis 
 triangle than in the upper. Of those infants who die in the first year of life,
 a proportion equal to $\lambda$ do not survive to December \nth{31} of the
 calendar year in which they were born.\footnote{$\lambda$ can be derived directly from death counts
 data classified by Lexis triangles. If the demographer
 does not have information to derive $\lambda$ directly, ad hoc or semidirect
 methods may be used to assign a reasonable proportion, such as 0.9 for
 contemporary low mortality populations.} The fertility component of
 $\textbf{Y}$, \textbf{$\Gamma$}, is then composed as in
 Matrix~\ref{matrix:Yfert}.

\begin{matrix}[h!]
\centering
\caption{Fertility component, \textbf{$\Gamma$}, of unisex thanatological
projection matrix.}
\label{matrix:Yfert}
$\bordermatrix{
  {e_y } \vspace{.6em}&                0_t  & 1_t  & 2_t  & 3_t  & 4_t  & 5_t\cr 
   0_{t+1} \vspace{.6em}& (1-\lambda) \tfrac{\gamma_0d_0}{2} & (1-\lambda) \gamma_1d_0 & (1-\lambda)
   \gamma_2d_0 & (1-\lambda) \gamma_3d_0 & (1-\lambda) \gamma_4d_0 & 0 \cr 
   1_{t+1} \vspace{.6em}& \tfrac{\gamma_0d_1}{2} & \gamma_1d_1 & \gamma_2d_1 & \gamma_3d_1 & \gamma_4d_1
   & 0   \cr 2_{t+1} \vspace{.6em}& \tfrac{\gamma_0d_2}{2} & \gamma_1d_2 & \gamma_2d_2 & \gamma_3d_2 & \gamma_4d_2
   & 0   \cr 3_{t+1} \vspace{.6em}& \tfrac{\gamma_0d_3}{2} & \gamma_1d_3 & \gamma_2d_3 & \gamma_3d_3 & \gamma_4d_3
   & 0   \cr 4_{t+1} \vspace{.6em}& \tfrac{\gamma_0d_4}{2} & \gamma_1d_4 & \gamma_2d_4 & \gamma_3d_4 & \gamma_4d_4
   & 0   \cr 5_{t+1} \vspace{.6em}& \tfrac{\gamma_0d_5}{2} & \gamma_1d_5 & \gamma_2d_5 & \gamma_3d_5 & \gamma_4d_5
   & 0   }$
\end{matrix}
\FloatBarrier

The survival matrix, \textbf{S}, and the fertility matrix, \textbf{$\Gamma$} sum
to create the thanatological projection matrix, \textbf{Y}:

\begin{equation}
\textbf{Y} = \textbf{S} + \textbf{$\Gamma$}
\end{equation}
